M.Sc. Proposal: \\ \bigskip Computational Electrokinetics

### 2012-08-26 notes/work/ch01.tex
Review.
Add electrophoresis theory to notes.

The partial differential equations that descibe the system dynamics
are coupled and non-linear. 
In general, they have no analytic solution. Moreover,
any numerical solver must handle the scale disparity caused by the
Debye layer width being much smaller than the particle's size. A
closed-form linear asymptotic solution has been developed for
spherical ion-exchanging particles and weak electric field in an
axisymmetric setting \cite{yariv2010migration} but it is hard to extend this
analytic solution to more general systems. Once the electric field
becomes stronger, significant nonlinear phenomena are expected.
This interesting regime has not yet been explored.

Recently, it has been conceived that such type of dynamics
can be very useful in transporting and manipulating micro-
and nanoscale objects for many nanotechnology applications,
such as the self-assembly of superstructures, roving sensors, 
drug-delivery systems and useful nanomachinery 
\cite{pumera2010electrochemically,paxton2004catalytic,howse2007self}.
However, since the analytic model for these electrokinetic
phenomena has no closed-form analytic solution, a numerical
solver has to be developed for analysing such system.

@@

Spatial coordinates $\br$ are normalized by $a^*$ (the characteristic size of the ion-exchanger).

@@

The dimensional ionic fluxes are given by Nernst-Plank equation 
(using Einstein relation for the electric mobility, $\mu_q^* = D^* / k_B^* T^*$):
\begin{eqnarray}
\bj^*_\pm &=& 
-D^* \bnabla c^*_\pm + \bv^* c^*_\pm \mp \frac{z e^* D^*}{k_B^* T^*} c^*_\pm \bnabla \varphi^*,
\end{eqnarray}
where $D^*_\pm \approx D^*$ is the ionic diffusivity.

@@

The dimensional Poisson equation is:
\begin{eqnarray}
(c^*_+ - c^*_-) z F^* &=& -\eps^* \Laplacian \varphi^*,
\end{eqnarray}
and after nondimensionalization, is:
\begin{eqnarray}
c_+ - c_- &=& -\frac{\eps^* \varphi^*}{z F^* c^* (a^*)^2} \Laplacian \varphi = 
-\delta^2 \Laplacian \varphi,
\\
\delta^2 &=& \frac{\eps^* \varphi^*}{c^* F^* z (a^*)^2} = 
\pars{\frac{\delta^*}{a^*}}^2.
\end{eqnarray}

@@

\subsubsection{Incompressible Stokes Flow with Electrostatic Force}
Mass conservation (due to constant fluid density):
\begin{eqnarray}
\bnabla \cdot \bv &=& 0.
\end{eqnarray}
Momentum conservation (force balance):
\begin{eqnarray}
\Laplacian \bv - \bnabla p + \bnabla \varphi \Laplacian \varphi &=& 0.
\end{eqnarray}
Free charge density is given by the Poisson equation:
\begin{eqnarray}
\delta^2 \Laplacian \varphi &=& -2q.
\end{eqnarray}

\subsubsection{Boundary conditions}
An constant electric field $\bE$ is applied to the system, such that $|\bE| = \beta$.
The ion-exchanger will acquire a non-zero steady-state velocity $\cU$, which is
a function of the applied field, and is not known a priori.

@@

We consider a spherical ion-exchanger. 
At its surface $\mathcal{S}$ (defined at $r=1$ with normal $\bn = \brhat$), 
we have zero slip, anion impermeability, cation selectivity (with fast Butler-Volmer kinetics)
and high conductance of the particle (yielding uniform electric potential):
\begin{eqnarray}
\bv & = & \bzero, \\
\bn \cdot \bj_- & = & 0, \\
\bn \cdot \bj_+ & = & k (1 - c_+/\gamma) \Rightarrow c_+ \rightarrow \gamma, \\
\varphi & = & \cV.
\end{eqnarray}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Separation of scales}
When $\delta \ll 1$, 
most of the fluid remains electroneutral ($q \approx 0$, thus $c_+ \approx c_-$), 
except a thin boundary layer of width $O(\delta)$, 
also known as the ``Debye layer'', surrounding the ion exchanger.

\subsubsection  {Bulk-scale equations}
The Nernst-Planck equations above for the fluid bulk (outside
the boundary layer), using bulk variables, read:
\begin{eqnarray}
  Q & = & 0, \\
  C & = & C_+ = C_-, \\
\bJ &=& -\bnabla C + \alpha \bV C, \\
\bI &=& -C \bnabla \varPhi.
\end{eqnarray}
Because the flow is incompressible ($\bnabla \cdot V = 0$), 
ion conservation results in the following equations:

@@

\subsubsection{Debye-scale effective boundary conditions}
We define the scaled coordinate $\rho$ (normal to ion-exchanger surface), where $\delta \ll 1$:
\begin{eqnarray}
  \rho &=& \frac{r-1}{\delta}. 
\end{eqnarray}
Thus, the electric potential and the ionic concentrations are $O(1)$ at the Debye layer ($r \approx 1$):
\begin{eqnarray}
  \varphi(\rho,\theta) &\sim& \varphi, \\
  c_\pm(\rho,\theta) &\sim& c_\pm, \\
  c(\rho,\theta) &\sim& c, \\
  q(\rho,\theta) &\sim& q.
\end{eqnarray}

@@

@@

>>>>

The partial differential equations that descibe the system dynamics
are coupled and non-linear. 
In general, they have no analytic solution. Moreover,
any numerical solver must handle the scale disparity caused by the
Debye layer width being much smaller than the particle's size. A
closed-form linear asymptotic solution has been developed for
spherical particles and weak electric field in an
axisymmetric setting \cite{yariv2010migration, schnitzer2012surface} but it is hard to extend this
analytic solution to more general systems. Once the electric field
becomes stronger, significant nonlinear phenomena are expected.
This interesting regime has not yet been explored.

Recently, it has been conceived that such type of dynamics
can be very useful in transporting and manipulating micro-
and nanoscale objects for many nanotechnology applications,
such as the self-assembly of superstructures, roving sensors, 
drug-delivery systems and useful nanomachinery 
\cite{howse2007self,paxton2004catalytic,pumera2010electrochemically}.
However, since the analytic model for these electrokinetic
phenomena has no closed-form analytic solution, a numerical
solver has to be developed for analysing such system.

@@

Spatial coordinates $\br$ are normalized by $a^*$ (the characteristic size of the particle).

@@

The dimensional ionic fluxes are given by Nernst-Plank equation 
(using Einstein relation for the electric mobility, $\mu_q^* = D^* / k_B^* T^*$):
\begin{eqnarray}
\bj^*_\pm &=& 
-D^* \bnabla c^*_\pm + \bv^* c^*_\pm \mp \frac{z e^* D^*}{k_B^* T^*} c^*_\pm \bnabla \varphi^*,
\end{eqnarray}
where $D^*$ is the ionic diffusivity.

@@

The dimensional Poisson equation is:
\begin{eqnarray}
(c^*_+ - c^*_-) z e^* &=& -\eps^* \Laplacian \varphi^*,
\end{eqnarray}
and after nondimensionalization, is:
\begin{eqnarray}
c_+ - c_- &=& -\frac{\eps^* \varphi^*}{z e^* c^* (a^*)^2} \Laplacian \varphi = 
-2\delta^2 \Laplacian \varphi,
\\
\delta^2 &=& \frac{\eps^* \varphi^*}{2c^* e^* z (a^*)^2} = 
\pars{\frac{\delta^*}{a^*}}^2.
\end{eqnarray}

@@

\subsubsection{Incompressible Stokes Flow with Electrostatic Force}
Mass conservation (due to constant fluid density):
\begin{eqnarray}
\bnabla \cdot \bv &=& 0.
\end{eqnarray}
Momentum conservation (force balance):
\begin{eqnarray}
\Laplacian \bv - \bnabla p + \bnabla \varphi \Laplacian \varphi &=& 0.
\end{eqnarray}
Free charge density is given by the Poisson equation:
\begin{eqnarray}
\delta^2 \Laplacian \varphi &=& -q.
\end{eqnarray}

\subsubsection{Boundary conditions}
An constant electric field $\bE$ is applied to the system, such that $|\bE| = \beta$.
The particle will acquire a non-zero steady-state velocity $\cU$, which is
a function of the applied field, and is not known a priori.

@@

We consider a spherical particle. 
At its surface $\mathcal{S}$ (defined at $r=1$ with normal $\bn = \brhat$), 
we have zero slip and ion impermeability.
\begin{eqnarray}
\bv & = & \bzero, \\
\bn \cdot \bj_\pm & = & 0.
\end{eqnarray}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Bulk-scale equations}
When $\delta \ll 1$, 
most of the fluid remains electroneutral ($q \approx 0$, thus $c_+ \approx c_-$), 
except a thin boundary layer of width $O(\delta)$, 
also known as the ``Debye layer'', surrounding the ion exchanger.

The Nernst-Planck equations above for the fluid bulk (outside
the boundary layer), using bulk variables, read:
\begin{eqnarray}
  Q & = & 0, \\
  C & = & C_+ = C_-, \\
\bJ &=& -\bnabla C + \alpha \bV C, \\
\bI &=& -C \bnabla \varPhi.
\end{eqnarray}
Because the flow is incompressible ($\bnabla \cdot V = 0$), 
ion conservation results in the following equations:

@@

\subsection{Debye-scale effective boundary conditions}
The effective boundary conditions depend on the specific physical and chemical properties 
of the particle. We consider the following scenarios:

\subsubsection{Inert particle}

We define the scaled coordinate $\rho$ (normal to ion-exchanger surface), where $\delta \ll 1$:
\begin{eqnarray}
  \rho &=& \frac{r-1}{\delta}. 
\end{eqnarray}
Thus, the electric potential and the ionic concentrations are $O(1)$ at the Debye layer ($r \approx 1$):
\begin{eqnarray}
  \varphi(\rho,\theta) &\sim& \varphi, \\
  c_\pm(\rho,\theta) &\sim& c_\pm, \\
  c(\rho,\theta) &\sim& c, \\
  q(\rho,\theta) &\sim& q.
\end{eqnarray}

@@

Due to ion impermeability and radial flux continuity:
\begin{eqnarray}
  \br \cdot \bj_\pm = -\deriv{c_\pm}{r} \mp c_\pm \deriv{\varPhi}{r} &=& 0, \\
-\deriv{C}{r} \mp C \deriv{\varPhi}{r} &=& 0, \\
\deriv{C}{r} = \deriv{\varPhi}{r} &=& 0.
\end{eqnarray}

\subsubsection{Ion-exchanger}

@@

\subsubsection{Highly charged particle}
\begin{eqnarray}
 \deriv{C}{r} + C \deriv{\varPhi}{r} &=& 0, \\
 \deriv{C}{r} - \mbox{Du} \Laplacian_s (\varPhi - \log C) &=& 0.
\end{eqnarray}
